Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-01 paper/paper.tex
implement proof-reading changes

\begin{abstract}
    Stochastic trace estimation is a family of well established techniques for approximating the trace of constant matrices. However, oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$. This is the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$. However, this approach does not scale well if $m$ is large. We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization for all values of the parameter $t$. We show that all of these estimators closely match their corresponding error guarantees for constant matrices. In particular, we show that for the Nyström++ method, we can guarantee an error of order $\varepsilon$ by just using $\mathcal{O}(\varepsilon^{-1})$ random vectors to build the estimate in all values of $t$, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has. We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method. Finally, we confirm our theoretical developments in multiple numerical experiments.
\end{abstract}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ with the Dirac delta distribution $\delta$. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a symmetric parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting. A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e. $\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}. However, in our setting we often cannot guarantee either of the two conditions. Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been taken for analyzing low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices we provide bounds on the approximation error for our method.

\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze a method for using the aforementioned estimators to estimate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation.

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{B}(t) = \begin{bmatrix}
        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n}
\end{equation}
where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
\begin{equation}
    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
    \label{equ:definition-trace}
\end{equation}
However, we assume that we only have access to products of this matrix with vectors for each $t \in [a, b]$, so this definition will not be directly useful for computing the trace.

@@

\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}. %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}. We bound its $L^1$ norm in terms of the integral over the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we aim to approximate.

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{- \sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

@@

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Following \cite{kressner-2023-randomized-lowrank}, we can show that the error $\lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F$ is measurable. Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

@@

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

%\todo{
%\begin{remark}
%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
%    \begin{equation}
%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
%    \end{equation}
%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
%\end{remark}
%}

@@

    The $p$-th moment of the second term can be processed with standard matrix norm manipulations to
    \begin{align}

>>>>

\begin{abstract}
    Stochastic trace estimators are a family of well established techniques for approximating the trace of constant matrices. Oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$. This is for example the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$. However, this approach does not scale well if $m$ is large. We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization reused for all values of the parameter $t$. We show that all of these estimators closely match their corresponding error guarantees for constant matrices. In particular, for the Nyström++ method, we can guarantee an error of at most $\varepsilon$ by computing the estimate with just $\mathcal{O}(\varepsilon^{-1})$ random vectors, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has. We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method. Finally, we confirm our theoretical developments in multiple numerical experiments.
\end{abstract}

@@

In numerous problems across physics, chemistry, engineering, and data science, how the eigenvalues of certain matrices are distributed can be indicative of what properties a system has: In electronic structure calculations eigenvalues represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be represented by the spectral density $\phi(t) = n^{-1} \sum_{i=1}^{n} \delta(t - \lambda_i)$ which places a rescaled Dirac delta distribution $\delta$ at each eigenvalue. Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been developed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature rule with the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. Further, because $\delta$ is hard to approximate and because in many applications, a rough estimate of the spectral density is already acceptable, a standard approach is to instead approximate the smoothed spectral density

@@

for a smoothing kernel $g_{\sigma}$, typically a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This translates the problem to computing the trace of a symmetric parameter-dependent matrix $\mtx{B}(t) \equiv g_{\sigma}(t \mtx{I}_n - \mtx{A}) \in \mathbb{R}^{n \times n}$. Directly assembling the matrix function requires diagonalizing the matrix $\mtx{A}$, which again is prohibitively expensive. Therefore, we estimate its trace implicitly by approximating products of the matrix function with vectors. For constant matrices, the most widespread methods for implicit trace estimation are the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and the variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making, chen-2023-krylovaware-stochastic}. Some of these estimators have already been successfully applied to parameter-dependent matrices \cite{lin-2017-randomized-estimation,chen-2023-krylovaware-stochastic}, but they have not yet been analyzed in this setting. A related technique is dynamic trace estimation for sequences $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ of matrices which satisfy $\lVert \mtx{B}(t_i) \rVert _F \leq 1$ and which change dynamically, i.e. $\lVert \mtx{B}(t_{i+1}) - \mtx{B}(t_i) \rVert _F \leq \alpha < 1$ for the Frobenius norm $\lVert \cdot \rVert _F$ and a small $\alpha$ \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace}. However, in our setting we often cannot guarantee either of the two conditions. Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been used in the analysis of low-rank approximations of parameter-dependent matrices \cite{kressner-2023-randomized-lowrank}.

\paragraph{New contributions} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

\paragraph{Structure} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter dependent matrices. We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.

@@

We consider parameter-dependent matrices of the form
\begin{equation}
    \mtx{B}(t) = \begin{bmatrix}
        b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
        b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
        \vdots & \vdots & \ddots & \vdots \\
        b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
    \end{bmatrix} \in \mathbb{R}^{n \times n}
\end{equation}
where $b_{ij}(t)$ are functions depending continuously on the parameter $t$ which takes values in the interval $[a,b]$. The trace of such a matrix is defined as
\begin{equation}
    \Trace(\mtx{B}(t)) = \sum_{i=1}^{n} b_{ii}(t).
    \label{equ:definition-trace}
\end{equation}
However, as is often the case for matrix functions or implicit matrices, we assume that we can only efficiently access this matrix through products with vectors for each $t \in [a, b]$, so this definition will not directly be useful for computing the trace.

@@

\paragraph{Nyström++ estimator} Finally, \cite{lin-2017-randomized-estimation} proposes an estimator which corrects for inaccuracies in the Nyström approximation \refequ{equ:nystrom-trace-estimator} by estimating the trace of its residual using the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
This is the parameter-dependent analogue of the Nyström++ estimator \cite{persson-2022-improved-variants}, which itself is based on the Hutch++ estimator \cite{meyer-2021-hutch-optimal}. %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

We first analyze the Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}. We bound its $L^1$-error in terms of the $L^1$-norm of the Frobenius norm $\lVert \cdot \rVert _F$ of the matrix whose trace we approximate.

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric and continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) \right| ~\mathrm{d}t \leq 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, 1)$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) | ~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

@@

    First, we consider $r(\mtx{B}(t), \vct{\psi})$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r(\mtx{B}, \vct{\psi})$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$ where $\lVert \cdot \rVert _2$ denotes the spectral norm. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right]
        &\leq k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Because $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$, we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(\mtx{B}, \vct{\psi})^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments $\mathbb{E}^{k}[\cdot] = \left(\mathbb{E}\left[ | \cdot |^{k} \right] \right)^{\sfrac{1}{k}}$ of $r(\mtx{B}, \vct{\psi})$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(\mtx{B}, \vct{\psi}) \right]
        \leq \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F \leq 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Following \cite{kressner-2023-randomized-lowrank}, we can show that $r(\mtx{B}(t), \vct{\psi})$ is measurable. Therefore, Minkowski's integral inequality \cite[theorem 202]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t  \right]
        \leq \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(\mtx{B}(t), \vct{\psi}) \right]~\mathrm{d}t
        \leq 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    Consequently, by Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(\mtx{B}(t), \vct{\psi})|~\mathrm{d}t \leq 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

@@

    %Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) %\rceil$, where $\lceil x \rceil$ is the ceiling function which returns the greatest %integer larger than or equal to $x$, we get that for all $\delta \in (0, e^{-\sfrac%{1}{2}})$
    %\begin{equation}
    %    k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\sfrac{1}{2\lceil \log%(\delta^{-1}) \rceil}}
    %    = \lceil \log(\delta^{-1}) \rceil e^{\sfrac{1}{2}\frac{\log(\delta^{-1})}%{\lceil \log(\delta^{-1}) \rceil}}
    %    \leq 2 \log(\delta^{-1}) e^{\sfrac{1}{2}},
    %\end{equation}
    %from which follows the second part of the theorem when $n_{\mtx{\Psi}} = (2^5 e^%{\sfrac{1}{2}} \varepsilon^{-1} \log(\delta^{-1}))^2$.

    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$, where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to $x$, we get that for all $\delta \in (0, 1)$ the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$, where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.

    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

%\begin{remark}
%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
%    \begin{equation}
%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
%    \end{equation}
%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
%\end{remark}

@@

    The $q$-th moment of the second term can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to
        \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
